---
title: "Survival Analysis of STC (plus some other) Apprenticeship Completions"
output: flexdashboard::flex_dashboard
runtime: shiny
resource_files:
- out/actual_plus_forecast.rds
- out/at_risk_plots.rds
- out/diagnosis.csv
- out/observed_vs_backcast.rds
- out/survival.rds
- out/dates.rds
- out/completions_vs_demand.rds
- out/new_regs_vs_demand.rds
- out/trade_noc_mapping.csv
---

```{r}
library(tidyverse)
library(here)
library(conflicted)
library(plotly)
library(formattable)
conflicts_prefer(dplyr::filter)
conflicts_prefer(formattable::style)
font_size=18
my_survplot <- function(mod, dat, grp) {
  plt <- survminer::ggsurvplot(mod,
    dat,
    pval = TRUE,
    censor = FALSE,
    conf.int = FALSE,
    ggtheme = theme_grey(),
    font.title=font_size,
    font.x=font_size,
    font.y=font_size,
    font.tickslab=font_size,
    font.legend=font_size
  ) +
    labs(
      title = grp,
      x = "Months since Registration",
      y = "Survival Probability"
    )
}
my_dt <- function(tbbl) {
  DT::datatable(tbbl,
                filter = 'top',
                extensions = "Buttons",
                rownames = FALSE,
                options = list(
                  columnDefs = list(list(className = "dt-center", targets = "_all")),
                  paging = TRUE,
                  scrollX = TRUE,
                  scrollY = TRUE,
                  searching = TRUE,
                  ordering = TRUE,
                  dom = "Btip",
                  buttons = list(
                    list(extend = "csv", filename = "some_file_name"),
                    list(extend = "excel", filename = "some_file_name")
                  ),
                  pageLength = 20
                )
  )
}

survival <- read_rds(here("out","survival.rds"))
observed_vs_backcast <- read_rds(here("out","observed_vs_backcast.rds"))
at_risk_plus_forecast <- read_rds(here("out","at_risk_plots.rds"))
actual_plus_forecast <- read_rds(here("out","actual_plus_forecast.rds"))
dates <- read_rds(here("out","dates.rds"))
completions_vs_demand <-  read_rds(here("out","completions_vs_demand.rds"))
new_regs_vs_demand <-  read_rds(here("out","new_regs_vs_demand.rds"))
```

Thumbs and Eyes
=======================================================

Inputs {.sidebar data-width=400}
-------------------------------------

* This initial page gives a broad overview of whether or not there will be enough new journey-persons and apprentices to meet demand.
* A green thumbs up indicates there will likely be a surplus.
* A red thumbs down indicates there will likely be a shortage.
* An orange eye indicates we should probably keep an eye on the situation. 
* More details on the methodology can be found in the other tabs.

Column
---------------------------------------------------------------------

```{r}
jp_risk <- completions_vs_demand|>
  group_by(NOC_Code_2021, NOC_2021, series)|>
  summarize(value=sum(value))|>
  pivot_wider(names_from = series, values_from = value)|>
  mutate(`Journeyperson Demand over Supply Ratio`=demand/supply)|>
  select(NOC_2021,`Journeyperson Demand over Supply Ratio`)
  
ap_risk <- new_regs_vs_demand|>
  group_by(NOC_Code_2021, NOC_2021, series)|>
  summarize(value=sum(value))|>
  pivot_wider(names_from = series, values_from = value)|>
  mutate(`Apprentice Demand over Supply Ratio`=demand/supply)|>
  select(NOC_2021, `Apprentice Demand over Supply Ratio`)
  

risk <- full_join(jp_risk, ap_risk)

formattable(risk, list(
  `Apprentice Demand over Supply Ratio` = formatter("span",
      style = x ~ case_when(x > 1.1 ~ "font-size:19px; color:red",
                            x < .8 ~ "font-size:19px; color:green",
                            TRUE ~ "font-size:19px; color:orange"),
      x ~ icontext(case_when(
         x  > 1.1 ~ "thumbs-down",
        x < .8 ~ "thumbs-up",
        TRUE ~ "eye-open"))),
  `Journeyperson Demand over Supply Ratio` = formatter("span",
       style = x ~ case_when(x > 1.1 ~ "font-size:19px; color:red",
                            x < .8 ~ "font-size:19px; color:green",
                            TRUE ~ "font-size:19px; color:orange"),
      x ~ icontext(case_when(
         x  > 1.1 ~ "thumbs-down",
        x < .8 ~ "thumbs-up",
        TRUE ~ "eye-open")))
  ))
```



Forecast of Apprentice Completions
=========================================================

Inputs {.sidebar data-width=400}
-------------------------------------

### Introduction:

* Registration for apprenticeship is a leading indicator of completion of apprenticeship.
* In this app we use survival analysis to uncover the dynamics governing the transition between registering for an apprenticeship and completion of apprenticeship: i.e. not everyone who registers completes, and the elapsed time between registration and completion varies across trade and time. 
* Specifically, we forecast the arrival rate of new registrants using exponential smoothing, and then apply the joint probabilities of completion to the arrival rate to get the expected number of completions.

### What makes a good forecast:

1) A reasonable forecast of arrivals (the rate of arrival of new registrants is predictable)
2) Survival curves that are stable over time: here we split the 24 years of completion data into 3 eras of 8 years each, in order to ascertain whether the transition dynamics are stable.
3) Backcasts (applying the transition probabilities to historic new registrant arrival data) are close the observed values of completions.

```{r}
selectInput(
  "trade",
  "Choose a trade:",
  sort(survival$trade_desc),
  sort(survival$trade_desc)[1] 
)
```

Column
-------------------------------------
    
### Arrival rate of new apprentice registrations
    
```{r, fig.retina=2}
renderPlotly({
  req(input$trade)
  plt <- at_risk_plus_forecast$plot[at_risk_plus_forecast$trade_desc==input$trade][[1]]
  plot <- plotly::ggplotly(plt)
  plot[['x']][['layout']][['shapes']] <- list(
    list(type = "rect",
         fillcolor = "black", line = list(color = "black"), opacity = 0.1,
         x0 = dates$x0, 
         x1 = dates$x1, 
         xref = "x",
         y0 = 0, 
         y1 = 1000, 
         yref = "y"))
  return(plot)
})
```
    
### Survival Curves

```{r, fig.retina=2}
renderPlot({
  req(input$trade)
  plt <- survival|>
    filter(trade_desc==input$trade)|>
    mutate(km_split_plot = pmap(list(km_split_model, data, trade_desc), my_survplot))
  plt$km_split_plot
})
```

Column
-------------------------------------

### Observed vs. Backcasts

```{r, fig.retina=2}
renderPlotly({
  req(input$trade)
  observed_vs_backcast$plot[observed_vs_backcast$trade_desc==input$trade][[1]]
})
```

### Forecasts of Apprentice Completions
    
```{r, fig.retina=2}
renderPlotly({
  req(input$trade)
  plt <- actual_plus_forecast$plot[actual_plus_forecast$trade_desc==input$trade][[1]]
  plot <- plotly::ggplotly(plt)
  plot[['x']][['layout']][['shapes']] <- list(
    list(type = "rect",
         fillcolor = "black", line = list(color = "black"), opacity = 0.1,
         x0 = dates$x0, 
         x1 = dates$x1, 
         xref = "x",
         y0 = 0, 
         y1 = 2000, 
         yref = "y"))
  return(plot)
})
```

Apprentice Completions vs. Job Openings
================================================================

Inputs {.sidebar data-width=250}
-------------------------------------

### Are we producing enough journey-persons to meet demand?

* Supply of journey-persons is the forecast of apprentice completions.
* Demand for journey-persons is assumed to be twice LMO replacement demand plus 1/3 of LMO expansion demand.
* Regarding replacement demand, we assume that for every retirement or death there is another journey-person that leaves the trade prematurely.
* Regarding expansion demand, the desired ratio of journey-person to apprentice is 1:2.
* Note that there is a many-to-one relationship between trades and occupations. 


Column
------------------------------------------------

### Supply vs. Demand of Journeypersons

```{r}
plt <- completions_vs_demand|>
  mutate(NOC_2021=str_trunc(NOC_2021, width=30))|>
  ggplot(aes(year, value, colour=series))+
  geom_line()+
  scale_x_continuous(breaks=seq(2024,2034,2))+
  scale_y_continuous(trans="log10", labels = scales::comma)+
  labs(x=NULL,y=NULL)+
  facet_wrap(~NOC_2021, nrow = 3)

plotly::ggplotly(plt)|>
  plotly::layout(legend = list(orientation = 'h', x = 0.45, y = 1.1))
```

New Apprentice Registrations vs. Job Openings
================================================================

Inputs {.sidebar data-width=250}
-------------------------------------

### Are there enough new apprentices to meet demand?

* Supply of apprentices is the forecast of new registrations. 
* The demand for apprentices is assumed to be twice that of new journey-persons: the desired ratio is 2:1.
* Note that there is a many-to-one relationship between trades and occupations. 

Column
------------------------------------------------

### Supply vs. Demand of New Apprentices

```{r}
plt <- new_regs_vs_demand|>
  mutate(NOC_2021=str_trunc(NOC_2021, width=30))|>
  ggplot(aes(year, value, colour=series))+
  geom_line()+
  scale_x_continuous(breaks=seq(2024, 2034,2))+
  scale_y_continuous(trans="log10", labels = scales::comma)+
  labs(x=NULL,y=NULL)+
  facet_wrap(~NOC_2021, nrow = 3)

plotly::ggplotly(plt)|>
  plotly::layout(legend = list(orientation = 'h', x = 0.45, y = 1.1))
```

Trades and occupations
========================================================

```{r}
my_dt(read_csv(here("out", "trade_noc_mapping.csv")))
```

